\section{Conclusion}
\label{sec:conclusion}

We systematically tested whether linear directions in the residual stream of \pythia compose under vector addition, evaluating 300 direction pairs across 25~concept categories with both geometric and functional metrics.
Our central finding is that the boundary of compositionality coincides with the boundary of linearity itself.
Morphological transformations form strong, compositional linear directions; encyclopedic associations form moderate ones; lexicographic semantic relations --- synonymy, antonymy, meronymy, hyponymy --- generally fail to form coherent linear directions at all, making the question of their compositionality moot.
Among well-formed directions, composition via vector addition is remarkably robust, with functional steering achieving $>$96\% accuracy across all pair types.

These results have direct implications for practitioners.
Steering vectors for well-defined transformations (verb tenses, pluralization, simple semantic associations) can be composed with confidence.
For abstract or relational concepts, practitioners should verify direction quality before attempting composition.
The Composition Fidelity Score and compositionality taxonomy we introduce provide tools for making this assessment.

Several directions remain open.
Testing on larger models would reveal whether lexicographic relations become more linear as capacity increases.
Extending the analysis to intermediate layers and behavioral concepts would broaden the scope of the compositionality map.
Finally, developing harder steering benchmarks that avoid the ceiling effects we observe would enable finer-grained comparisons between pair types.
